Component response rate variation drives stability in large complex systems

### 2018-07-26 notebook/ms_for_review.Rmd
Add more edits

Rows in $M$ define how a given component $i$ is affected by other components of the system, meaning that the rate of component response time can be modelled by multiplying all row elements by a real scalar value $\gamma_{i}$ [@Patel2018] (see Supplementary Information for details). The distribution of $\gamma$ over $S$ components thereby models the distribution of component response rates. An instructive example compares one $M$ where $\gamma_{i} = 1$ for all $i$ in $S$ to the same $M$ when half of $\gamma_{i} = 1.95$ and half of $\gamma_{i} = 0.05$. This models one system in which $\gamma$ is invariant and one in which $\gamma$ varies, but systems are otherwise identical (note $E[\gamma_{i}] = 1$ in both cases). I assume $S = 200$, $C = 0.05$, and $\sigma = 0.4$; diagonal elements are set to $-1$ and non-zero off-diagonal elements are drawn randomly from $\mathcal{N}(0, \sigma^{2})$. Rows are then multiplied by $\gamma_{i}$ to generate $M$. When $\gamma_{i} = 1$, eigenvalues of $M$ are distributed uniformly within a circle centred at ($-1, 0$) with a radius of `r round(sigma*sqrt(S*C), digits = 3)` (Fig. 1a). Hence, the real components of eigenvalues are highly unlikely to all be negative when all $\gamma_{i} = 1$. But when $\gamma_{i}$ values are separated into two groups, eigenvalues are no longer uniformly distributed (Fig. 1b). Instead, two distinct clusters of eigenvalues appear (grey circles in Fig. 1b), one centred at ($-1.95, 0$) and the other centred at ($-0.05, 0$). The former has a large radius, but the real components have shifted to the left (in comparison to when $\gamma = 1$) and all $\Re({\lambda}) < 0$. The latter cluster has real components that have shifted to the right, but has a smaller radius. Overall, for 1 million randomly assembled $M$, this division between slow and fast component response rates results in more stable systems: `r sum(pr_st[,1])` stable given $\gamma = 1$ versus `r sum(pr_st[,2])` stable given $\gamma = \{1.95, 0.5\}$.

@@

Higher stability in systems with variation in $\gamma$ can be observed by sampling $\gamma_{i}$ values from various distributions. I now focus on a uniform distribution where $\gamma \sim \mathcal{U}(0, 2)$ (see Supplementary Information for other distributions, which give similar results). As with the case of $\gamma = \{1.95, 0.5\}$ (Fig. 1b), $E[\gamma] = 1$ when $\gamma \sim \mathcal{U}(0, 2)$, allowing comparison of $M$ before and after variation in component response rate. Figure 2 shows a comparison of eigenvalue distributions given $S = 1000$, $C = 0.05$, and $\sigma = 0.4$. As expected [@Tao2010], when $\gamma = 1$, eigenvalues are distributed uniformly in a circle centred at ($-1, 0$) with a radius of $\sigma\sqrt{SC} =$ `r round(sigma*sqrt(S*C), digits = 3)`. Uniform variation in $\gamma$ leads to a non-uniform distribution of eigenvalues, some of which are clustered locally near the centre of the distribution, but others of which are spread outside the former radius of `r round(sigma*sqrt(S*C), digits = 3)` (Fig 2b). The clustering and spreading of eigenvalues introduced by $Var(\gamma)$ can destabilise previously stable systems or stabilise systems that are otherwise unstable. But where systems are otherwise too complex to be stable given $\gamma = 1$, the effect of $Var(\gamma)$ can often lead to stability above May's [@May1972; @Allesina2012] threshold $\sigma\sqrt{SC} < 1$.

@@

To investigate the effect of $Var(\gamma)$ on system stability, I simulated random $M$ matrices at $\sigma = 0.4$ and $C = 1$ across $S$ ranging from 2-32. One million $M$ were simulated for each $S$, and the stability of $M$ was assessed given $\gamma = 1$ versus $\gamma \sim \mathcal{U}(0, 2)$. For all $S > 10$, I found that the number of stable random systems was higher given $Var(\gamma)$ than when $\gamma = 1$ (Fig. 3), and that the difference between the probabilities of observing a stable system increased with an increase in $S$; i.e., the potential for $Var(\gamma)$ to drive stability increased with system complexity. For the highest values of $S$, nearly all systems that were stable given $Var(\gamma)$ would not have been stable given $\gamma = 1$ (see Supplementary Information for full results). This suggests that the stability of large systems might be dependent upon variation in the response rate of their individual components, meaning that factors such as generation time (in ecological networks), transaction speed (in economic networks), or communication speed (in social networks) needs to be considered when investigating the stability of complex systems.

@@

To further investigate the potential of $Var(\gamma)$ to be stabilising, I used a genetic algorithm (the space of possible $\gamma$ values was too large to search exhaustively [@Hamblin2013]; see Supplementary Information). For each of 40000 random $M$ ($\sigma = 0.4$, $C = 1$), the genetic algorithm initialised 1000 different sets of $\gamma \sim \mathcal{U}(0, 2)$ values of size $S$. Eigenanalysis was performed on $M$ using each set of $\gamma$ values, and the 20 sets with the lowest $\max\left(\Re(\lambda)\right)$ each produced 50 clonal offspring with subsequent mutation and crossover between the resulting new population of 1000 $\gamma$ sets. The genetic algorithm terminated if a stable $M$ was found, 20 generations occurred, or a convergence criteria of minimum fitness increase between generations was satisfied. Across $S = \{2, 3, ..., 39, 40\}$, sets of $\gamma$ values were found that resulted in stable systems with probabilities that were up to four orders of magnitude higher than when $\gamma = 1$ (see Supplementary Information), meaning that stability could often be achieved by manipulating $S$ $\gamma$ values rather than $S \times S$ $M$ elements. Hence, managing the response rates of system components in a targetted way can potentially facilitate the stabilisation of complex systems through a reduction in dimensionality.

@@

@@

>>>>

**Component response rates of random complex systems**. Rows in $M$ define how a given component $i$ is affected by other components of the system, meaning that the rate of component response time can be modelled by multiplying all row elements by a real scalar value $\gamma_{i}$ [@Patel2018]. The distribution of $\gamma$ over $S$ components thereby models the distribution of component response rates. An instructive example compares one $M$ where $\gamma_{i} = 1$ for all $i$ in $S$ to the same $M$ when half of $\gamma_{i} = 1.95$ and half of $\gamma_{i} = 0.05$. This models one system in which $\gamma$ is invariant and one in which $\gamma$ varies, but systems are otherwise identical (note $E[\gamma_{i}] = 1$ in both cases). I assume $S = 200$, $C = 0.05$, and $\sigma = 0.4$; diagonal elements are set to $-1$ and non-zero off-diagonal elements are drawn randomly from $\mathcal{N}(0, \sigma^{2})$. Rows are then multiplied by $\gamma_{i}$ to generate $M$. When $\gamma_{i} = 1$, eigenvalues of $M$ are distributed uniformly within a circle centred at ($-1, 0$) with a radius of `r round(sigma*sqrt(S*C), digits = 3)` (Fig. 1a). Hence, the real components of eigenvalues are highly unlikely to all be negative when all $\gamma_{i} = 1$. But when $\gamma_{i}$ values are separated into two groups, eigenvalues are no longer uniformly distributed (Fig. 1b). Instead, two distinct clusters of eigenvalues appear (red circles in Fig. 1b), one centred at ($-1.95, 0$) and the other centred at ($-0.05, 0$). The former has a large radius, but the real components have shifted to the left (in comparison to when $\gamma = 1$) and all $\Re({\lambda}) < 0$. The latter cluster has real components that have shifted to the right, but has a smaller radius. Overall, for 1 million randomly assembled $M$, this division between slow and fast component response rates results in more stable systems: `r sum(pr_st[,1])` stable given $\gamma = 1$ versus `r sum(pr_st[,2])` stable given $\gamma = \{1.95, 0.5\}$.

@@

Higher stability in systems with variation in $\gamma$ can be observed by sampling $\gamma_{i}$ values from various distributions. I focus on a uniform distribution where $\gamma \sim \mathcal{U}(0, 2)$ (see Supplementary Information for other distributions, which give similar results). As with the case of $\gamma = \{1.95, 0.5\}$ (Fig. 1b), $E[\gamma] = 1$ when $\gamma \sim \mathcal{U}(0, 2)$, allowing comparison of $M$ before and after variation in component response rate. Figure 2 shows a comparison of eigenvalue distributions given $S = 1000$, $C = 0.05$, and $\sigma = 0.4$. As expected [@Tao2010], when $\gamma = 1$, eigenvalues are distributed uniformly in a circle centred at ($-1, 0$) with a radius of $\sigma\sqrt{SC} =$ `r round(sigma*sqrt(S*C), digits = 3)`. Uniform variation in $\gamma$ leads to a non-uniform distribution of eigenvalues, some of which are clustered locally near the centre of the distribution, but others of which are spread outside the former radius of `r round(sigma*sqrt(S*C), digits = 3)` (Fig 2b). The clustering and spreading of eigenvalues introduced by $Var(\gamma)$ can destabilise previously stable systems or stabilise systems that are otherwise unstable. But where systems are otherwise too complex to be stable given $\gamma = 1$, the effect of $Var(\gamma)$ can often lead to stability above May's [@May1972; @Allesina2012] threshold $\sigma\sqrt{SC} < 1$.

@@

**Simulation of random $\mathbf{M}$ across $\mathbf{S}$**. To investigate the effect of $Var(\gamma)$ on system stability, I simulated random $M$ matrices at $\sigma = 0.4$ and $C = 1$ across $S$ ranging from 2-32. One million $M$ were simulated for each $S$, and the stability of $M$ was assessed given $\gamma = 1$ versus $\gamma \sim \mathcal{U}(0, 2)$. For all $S > 10$, I found that the number of stable random systems was higher given $Var(\gamma)$ than when $\gamma = 1$ (Fig. 3), and that the difference between the probabilities of observing a stable system increased with an increase in $S$; i.e., the potential for $Var(\gamma)$ to drive stability increased with system complexity. For the highest values of $S$, nearly all systems that were stable given $Var(\gamma)$ would not have been stable given $\gamma = 1$ (see Supplementary Information for full results).

**Targetted manipulation of $\mathbf{\gamma}$**. To further investigate the potential of $Var(\gamma)$ to be stabilising, I used a genetic algorithm (the space of possible $\gamma$ values was too large to search exhaustively [@Hamblin2013]). [Need a methods like sentence here](http://google.com). Across $S = \{2, 3, ..., 39, 40\}$, sets of $\gamma$ values were found that resulted in stable systems with probabilities that were up to four orders of magnitude higher than when $\gamma = 1$ (Fig. 4), meaning that stability could often be achieved by manipulating $S$ $\gamma$ values rather than $S \times S$ $M$ elements.

This suggests that the stability of large systems might be dependent upon variation in the response rate of their individual components, meaning that factors such as generation time (in ecological networks), transaction speed (in economic networks), or communication speed (in social networks) needs to be considered when investigating the stability of complex systems.

@@

@@

Hence, managing the response rates of system components in a targetted way can potentially facilitate the stabilisation of complex systems through a reduction in dimensionality.

@@

For each of 40000 random $M$ ($\sigma = 0.4$, $C = 1$), the genetic algorithm initialised 1000 different sets of $\gamma \sim \mathcal{U}(0, 2)$ values of size $S$. Eigenanalysis was performed on $M$ using each set of $\gamma$ values, and the 20 sets with the lowest $\max\left(\Re(\lambda)\right)$ each produced 50 clonal offspring with subsequent mutation and crossover between the resulting new population of 1000 $\gamma$ sets. The genetic algorithm terminated if a stable $M$ was found, 20 generations occurred, or a convergence criteria of minimum fitness increase between generations was satisfied.
